---
title: "Intro to Data Analysis Demonstration Data"
author: "Marney Pratt"
date: "October 1, 2025"
format: 
  html:
    code-overflow: wrap
    embed-resources: true
editor_options: 
  chunk_output_type: inline
---

## Load necessary packages

Run the code below at the start of the session to make sure all the necessary packages are loaded.

It is recommended that you put all your code to load packages in one R chunk at the start. Run the chunk below by clicking the little green arrow at the top right of the code chunk.

```{r}
#| label: setup
#| warning: false
#| message: false

suppressPackageStartupMessages(
  library(tidyverse)) ## for readr, dplyr, ggplot2
library(ggbeeswarm) ## for jittering points (you can remove this is you did not use a dotplot)
library(flextable) ## for making formatted tables that can be pasted into Word or Google Docs
library(dabestr) ##for data analysis using Bootstrap-coupled estimation
library(rstatix) #pipe friendly framework for basic stats
library(ggpubr) #publication ready plots
library(labelled) #for labeling variables better
library(glue) #for adding results to graph


```

## Demo Data

Below are some randomly generated (NOT REAL) data to use for practice. The code for generating the demo data and doing the estimation statistics is based on the [basics tutorial from the dabestr package](https://cran.r-project.org/web/packages/dabestr/vignettes/tutorial_basics.html). Run the following code chunk as is (you don't need to change anything)

```{r}
#| label: demo data

# sets some conditions
set.seed(12345) # Fix the seed so the results are replicable.
N <- 10 # The number of samples taken from each population

# Create samples for three different treatments
# each sample has a normal distribution 
c1 <- rnorm(N, mean = 3, sd = 0.4)
t1 <- rnorm(N, mean = 4, sd = 0.5)
t2 <- rnorm(N, mean = 3.2, sd = 0.6)


# Add an `id` column for paired data plotting.
id <- 1:N

# Combine samples into a DataFrame.
df <- tibble::tibble(
  `Baseline` = c1, 
  `Exercise` = t1, 
  `Recovery` = t2, 
  ID = id)

df <- df %>%
  tidyr::gather(key = Group, value = Measurement, -ID)


df <- df |> 
  labelled::set_variable_labels(
    ID                = "Subject ID",
    Group             = "Time of Measurement",
    Measurement       = "Natural Logarthim of Root Mean Square of Successive Differences"
  )

#see https://www.pipinghotdata.com/posts/2022-09-13-the-case-for-variable-labels-in-r/ for more help with labeling data

#make the catagorical variable a factor
df$Group = factor(df$Group,
                    levels = c("Baseline","Exercise","Recovery"))


```

Take a look at the dataset we just created by clicking on "df" in the Environment tab in the upper right RStudio window.

Note that this dataset, which we named "df" for data frame, includes a tidy data spreadsheet. There are three columns and 30 observations in each. Each column has a column name and a more descriptive label that we added using the `set_variable_labels` function from the `labelled` package.

The randomly generated data is set up to be similar to the experiment we did earlier. In this case, there are 10 subjects (because we set N<- 10). Each subject had a HRV measure (in this case, Natural Logarithm of Root Mean Square of Successive Differences or ln RMSSD) recorded during a 5 min baseline recording, during 5 minutes of exercise, and 5 minutes of recovery.

## Define Variables

To make using template code easier and avoid typing the variable names many times, we will define our variables below to change the column names used in the dataframe to be more standardize (for example the "Measurement" column that contains the continuous variable of interest is relabeled as "cont.var" )

```{r}
#| label: define variables

# replace the blank below with a continuous variable you want to summarize
cont.var <- "Measurement" 

# replace the blank below with the categorical variable (also called a factor) 
# this will be used to group the continuous variable
factor.var <- "Group" 

```


## Descriptive Statistics

First, let's get the full descriptive statistics by running the next code chunk...

```{r}
#| label: descriptive stats

# the code below will calculate descriptive statistics for a variable of interest grouped by another variable

# create a new dataframe called 'df.sum'
# start with the full dataframe (in this case, called 'df') then..
df.sum <- df  |> 
  
  # remove missing values from continuous variables
  filter(!is.na(.data[[cont.var]])) |> 
  
  # Group the data by your categorical variable
  group_by(.data[[factor.var]]) |> 
  
  # calculate the descriptive stats
  summarize(Sample.size = n(),
            Min = min(.data[[cont.var]]),
            Q1 = quantile(.data[[cont.var]], .25),
            Median = median(.data[[cont.var]]),
            Q3 = quantile(.data[[cont.var]], .75),
            Max = max(.data[[cont.var]]),
            Mean = mean(.data[[cont.var]]),              
            SD = sd(.data[[cont.var]]))

df.sum



```

How should we round these variables? A good rule of thumb is to use the SD as the indicator of how much average variability there is in your data. Personally, I like to round the SD to 2 significant digits based on recommendations in [Cole et al. (2015)](https://www-ncbi-nlm-nih-gov.libproxy.smith.edu/pmc/articles/PMC4483789/). Then round all the other values to the same decimal place as the rounded SD.

In the case above, we should round the SD to the hundredths place to get it rounded to 2 significant digits. Thus, we will round all the values to the hundredths place.

We also don't need to show ALL these values (in fact, it may not be important to give any of these exact values if the graph itself is sufficient to get your point across). To know what values to report as well as how to analyze and graph the results, we need to see if the data has a normal distribution and if there are any outliers.

### Test Assumption

First we will use a QQ plot to see if the distribution is normal

```{r}
#| label: normal distribution

#make a QQ plot for the continuous variable
#this tests to see if each group has a normal distribution
ggpubr::ggqqplot(
  data = df,                  #put the data frame here 
  x = {{ cont.var }},            #put the continuous variable here
  facet.by = {{ factor.var }}  #put the factor variable here
)  

```


Above shows a Q-Q plot (quantile-quantile plot) for each group comparing to a normal distribution in a graphical manner (see this [Complete Guide: How to Interpret Q-Q Plots](https://www.statology.org/qq-plot-interpretation/) if you want to learn more). Because the points are mostly close to the line and don't stray much outside of the grey shaded area, this indicates that the data are pretty normally distributed.

Next, we will look for outliers.

```{r}
#| label: test for outliers

#test for outliers
df  |> 
  group_by(.data[[factor.var]])  |> 
  rstatix::identify_outliers({{ cont.var }}) 

```


The result above shows how many rows contain outliers. It says 0 rows, which indicates that there were no outliers.

### Reporting Descriptive Statistics

The results above indicate that this dataset has a normal distribution, so it would be appropriate to give the mean, SD, and number of subjects (N). The code below would round to the hundredths place and give just the values of interest in a nice formatted table (using the `flextable` package to format the table).

```{r}
#| label: formatted mean SD

#see https://davidgohel.github.io/flextable/ for more information and formatting options


#calculate descriptive stats
df.sum <- df |> 
  group_by(.data[[factor.var]]) |> # grouping variable
  filter(!is.na(.data[[cont.var]])) |> # remove missing values 
  
  #calculate the rounded values
  summarise(Mean = round(mean(.data[[cont.var]]), digits=2), 
            SD = signif(sd(.data[[cont.var]]), digits=2),
            N = n())


#create the formatted table
ft <- flextable(df.sum,
                cwidth = 0.75) |>  #can vary cell width as needed
  
  #bold the headings
  bold(part = "header") |> 
  
  #center columns
  align(align = "center", part = "all" )

#print the table
#right click on the table, choose select all, 
#choose copy, then paste in your document
#finish formatting as needed in your document
ft
```

If this dataset was not a normal distribution, you might choose to report the median, range (min - max), and N.

```{r}
#| label: formatted descriptive stats

#see https://davidgohel.github.io/flextable/ for more information and formatting options


#calculate descriptive stats
df.med <- df |> 
  group_by(.data[[factor.var]]) |> #grouping variable
  filter(!is.na(.data[[cont.var]])) |> # remove missing values 
  
  summarise(Median = round(median(.data[[cont.var]]), digits=2),
            min = round(min(.data[[cont.var]]), digits=2),
            max = round(max(.data[[cont.var]]), digits=2),
            N = n()) |> 
  mutate(Range = paste(min, max, sep = "-")) |> 
  dplyr::select(-min, -max) |> 
  relocate(N, .after = last_col())
  


#create the table
ft <- flextable(df.med,
                cwidth = 0.75) |>  #can vary cell width as needed
  
  #bold the headings
  bold(part = "header") |> 
  
  #center columns
  align(align = "center", part = "all" ) |> 
  
  set_table_properties(layout = "autofit")

#print the table
#right click on the table, choose select all, 
#choose copy, then paste in your document
#finish formatting as needed in your document
ft



```
 In general, you should choose the most appropriate way to report your descriptive statistics based on the sample size and distribution of the data. Don't give ALL the districptive statistics  - I would pick one of the options above, not both. In this case, since the data show a normal distribution, I would report the Mean, SD, and N.


## Graphing the Results

Next we will look at the data to see if we can discern any interesting patterns. Since the sample size is small and we have repeated measures for the same subjects, we will show all the points and connect them for each subject. We will also include a measure of central tendency for all the subects at each time point.

```{r}
#| label: graph without stats

#the code below creates a plot that connects the dots within subjects

line.plot <- df  |>    
  ggplot(aes(
    x=Group, 
    y=Measurement, 
    group = `ID`)) +
  geom_point(size = 3,  
             color = "orange", 
             alpha = 0.5) +
  geom_line(linewidth = 0.2, 
            alpha = 0.5) + 
  stat_summary(fun = mean, 
               fun.min = mean, 
               fun.max = mean, 
               geom = "crossbar", 
               color = "black",
               width = 0.2, 
               size = 0.5,
      aes(group = 'Group' )) +
  xlab("Time of Measurement") +
  ylab("ln RMSSD") +

  #additional formatting
  theme_classic(base_size = 14)  +  #sets the font size
   theme(
    plot.title = element_text(size = 12),  # Set title size
    plot.subtitle = element_text(size = 12) # Set subtitle size to the same value
  ) +
  theme(legend.position = "none")   #controls legend/key

line.plot


```

Please note that it is important to include units on axes. The x-axis has a category without units. The y-axis in this case also has no units. While RMSSD does have units of milliseconds, when you take the natural logarithm it becomes dimensionless.

It looks like exercise increased the ln RMSSD and then it went back down to something similar to the baseline during recovery. But how big is the difference?


## Effect Size

An effect size tells us how big a difference is and whether that difference is significantly different from zero.  

How do we know if there is a difference in the HRV value during exercise? We use inferential statistics to help us determine whether a sample can help tell us about the full population and make comparisons. In this case, we want to compare between two groups at a time: baseline vs exercise to tell us if exercise had an impact, and baseline vs recovery to see if ln RMSSD returned to the baseline value.

You will learn much more about inferential statistics in a full statistics course. In this class, we are just going to discuss effect sizes. An effect size tells us how big the difference is between groups. A standardized effect size takes the difference between groups but also divides that difference by some measure of variability.

To pick the right test, we need to determine if the distribution is normal and if the variance is equal.

-   **Normal Distribution, Equal Variance:** If the data is normal, and the two groups have similar dispersion (i.e. similar variance), then you should use "regular" **Cohen's d effect size (*d*)**

-   **Unequal Variance:** If the data has unequal dispersion (i.e. unequal variance), then you should use the **Cohen's d for Welch test effect size (*d*)**

-   **Non-normal Distribution, Equal Variance:** If the data is NOT normal but the two groups have similar dispersion (i.e. similar variance), then you should use the **Wilcoxon effect size (*r*)** (note that some statisticians recommend to use the Cohen's d for Welch test effect size if the variance is very different even if the distribution is not normal)

We already tested the distribution above and found that it was normal for this demo dataset. Note that the variance is the square of the standard deviation. So just look at the SD values in the full descriptive statistics table to see if the SD is fairly similar. Since the SD for the recovery time period was almost twice as large as the baseline, we used the **Cohen's d for Welch test effect size (*d*)** below.

```{r}
#| label: cohen's d effect size


#set the formula to use the defined variables 
#this makes it easier to change the variables at the top and rerun everything 
#without having to retype the variables in every code chunk
    response <- cont.var
    predictor <- factor.var
    formula_string <- paste0(response, " ~ ", predictor)
    formula_object <- as.formula(formula_string)

#standardized effect size for normal data with unequal variance
#Cohen's d
effectSize.results  <- 
  rstatix::cohens_d(
    data = df,           #put the dataframe name here
    formula_object,      #continuous variable ~ factor (see above)
    paired = TRUE,       #TRUE for within-subjects, FALSE for between-subjects
    var.equal = FALSE,   #TRUE for equal variance, FALSE for unequal variance
    hedges.correction = TRUE,
    ci = TRUE,
    conf.level = 0.95,
    ci.type = "bca",
    nboot = 1000)

effectSize.results <- effectSize.results |> 
 mutate(comparison = paste(group1, group2, sep = "-")) 

effectSize.results



```

The effect size is the difference between typical values standardized by the amount of variation in the data (the exact equation depends on which effect size you use). The effect size here was calculated by taking 1000 bootstrapped random samples and the 95% confidence intervals were calculated with the adjusted bootstrap percentile (BCa) method.

The interpretation of the standardized effect size is included in the magnitude column - negligible, small, moderate, or large. This will help you determine if the difference between means or medians is meaningful or not. The 95% confidence intervals tells you how precisely the effect size was estimated. If the 95% confidence interval does not include zero, it suggests that the effect is statistically significant

The Baseline - Exercise comparison had a large magnitude effect size of -1.8 (the value is negative because it subtracted Exercise from the Baseline and Exercise was greater, also note that we usually round effect sizes to 2 significant figures). The 95% Confidence interval values were also both negative, which means that the 95% CI does not cross over zero - indicating that the average ln RMSSD value was significantly greater after exercise than at baseline. 

The Baseline - Recovery comparison had a small magnitude effect size of -0.33. The lower 95% CI was negative (below zero), but the upper 95% CI was positive (above zero). This indicates that the 95% CI crosses zero and means that the average ln RMSSD value was not significantly different from baseline.

::: {.callout-note appearance="minimal"}
## Optional Additional Resources on Effect Sizes

-   How to use the `rstatix` package to calculate effect sizes

    -   **Cohen’s d for Student t-test** on the [T-test Effect Size using Cohen's d Measure](https://www.datanovia.com/en/lessons/t-test-effect-size-using-cohens-d-measure/) webpage

    -   **Cohen’s d for Welch test** on the [T-test Effect Size using Cohen's d Measure](https://www.datanovia.com/en/lessons/t-test-effect-size-using-cohens-d-measure/) webpage

    -   **Effect size** under **Wilcoxon rank sum test** on the [Wilcoxon Test in R](https://www.datanovia.com/en/lessons/wilcoxon-test-in-r/) webpage

-   To learn more about inferential tests to compare two groups

    -   compare means with normal distribution: [t-Test](https://datatab.net/tutorial/t-test)

    -   rank-sum test to compare two independent samples that are not normally distributed: [Mann-Whitney U-Test](https://datatab.net/tutorial/mann-whitney-u-test)

    -   signed-rank test to compare paired groups that are not normally distributed: [Wilcoxon signed-rank test](https://datatab.net/tutorial/wilcoxon-test)

-   [Alternatives to P value: confidence interval and effect size](https://pmc.ncbi.nlm.nih.gov/articles/PMC5133225/)
:::


Let's make a graph of the effect size results to make it more visual 


```{r}


#the code below creates a plot that connects the dots within subjects
#and also adds the Cohen's d results from the analysis above to the graph
effect.plot <- effectSize.results  |>    
  ggplot(aes(
    x=comparison, 
    y=effsize)) +
  
  geom_pointrange(aes(ymin = conf.low, ymax = conf.high),
                  size = 1, # Adjust point size
                  linewidth = 0.8, # Adjust line thickness
                  color = "steelblue") + # Set color
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") + # Add a reference line at 0
  xlab("Comparison") +
  ylab("Standardized Effect Size") +
  #additional formatting
  theme_classic(base_size = 14)  +  #sets the font size
   theme(
    plot.title = element_text(size = 12),  # Set title size
    plot.subtitle = element_text(size = 12) # Set subtitle size to the same value
  ) +
  theme(legend.position = "none")   #controls legend/key

effect.plot

```

This shows the effect size results visually. In this case, the points represent the value of the Cohen's d for Welch test for three comparisons, and the vertical lines represent the 95% confidence interval calculated using the adjusted bootstrap percentile (BCa) method.
If the 95% CI includes zero (shown by crossing over the dotted line), then that effect size is not significantly different from zero (meaning there is no evidence of an effect). Thus, this indicates that exercise had a large, significant effect on ln RMSSD. This also indicates that the ln RMSSD returned to values similar to baseline during recovery (there was only a small but NOT significant effect when comparing baseline and recovery).
