---
title: "Intro to Data Analysis Demonstration Data"
author: "Marney Pratt"
date: "October 1, 2025"
format: 
  html:
    code-overflow: wrap
    embed-resources: true
editor_options: 
  chunk_output_type: inline
---

## Load necessary packages

Run the code below at the start of the session to make sure all the necessary packages are loaded.

It is recommended that you put all your code to load packages in one R chunk at the start. Run the chunk below by clicking the little green arrow at the top right of the code chunk.

```{r}
#| label: setup
#| warning: false
#| message: false

suppressPackageStartupMessages(
  library(tidyverse)) ## for readr, dplyr, ggplot2
library(ggbeeswarm) ## for jittering points (you can remove this is you did not use a dotplot)
library(flextable) ## for making formatted tables that can be pasted into Word or Google Docs
library(dabestr) ##for data analysis using Bootstrap-coupled estimation
library(rstatix) #pipe friendly framework for basic stats
library(ggpubr) #publication ready plots
library(labelled) #for labeling variables better
library(glue) #for adding results to graph


```

## Demo Data

Below are some randomly generated (NOT REAL) data to use for practice. The code for generating the demo data and doing the estimation statistics is based on the [basics tutorial from the dabestr package](https://cran.r-project.org/web/packages/dabestr/vignettes/tutorial_basics.html). Run the following code chunk as is (you don't need to change anything)

```{r}
#| label: demo data

# sets some conditions
set.seed(12345) # Fix the seed so the results are replicable.
N <- 10 # The number of samples taken from each population

# Create samples for three different treatments
# each sample has a normal distribution 
c1 <- rnorm(N, mean = 3, sd = 0.4)
t1 <- rnorm(N, mean = 4, sd = 0.5)
t2 <- rnorm(N, mean = 3.2, sd = 0.6)


# Add an `id` column for paired data plotting.
id <- 1:N

# Combine samples into a DataFrame.
df <- tibble::tibble(
  `Baseline` = c1, 
  `Exercise` = t1, 
  `Recovery` = t2, 
  ID = id)

df <- df %>%
  tidyr::gather(key = Group, value = Measurement, -ID)


df <- df |> 
  labelled::set_variable_labels(
    ID                = "Subject ID",
    Group             = "Time of Measurement",
    Measurement       = "Natural Logarthim of Root Mean Square of Successive Differences"
  )

#see https://www.pipinghotdata.com/posts/2022-09-13-the-case-for-variable-labels-in-r/ for more help with labeling data

```

Take a look at the dataset we just created by clicking on "df" in the Environment tab in the upper right RStudio window.

Note that this dataset, which we named "df" for data frame, includes a tidy data spreadsheet. There are three columns and 30 observations in each. Each column has a column name and a more descriptive label that we added using the `set_variable_labels` function from the `labelled` package.

The randomly generated data is set up to be similar to the experiment we did earlier. In this case, there are 10 subjects (because we set N<- 10). Each subject had a HRV measure (in this case, Natural Logarithm of Root Mean Square of Successive Differences or ln RMSSD) recorded during a 5 min baseline recording, during 5 minutes of exercise, and 5 minutes of recovery.

## Descriptive Statistics

First, let's get the full descriptive statistics by running the next code chunk...

```{r}
#| label: descriptive stats

# the code below will calculate descriptive statistics for a variable of interest grouped by another variable

# replace the blank below with a continuous variable you want to summarize
dep.var <- "Measurement" 

# create a new dataframe called 'df.sum'
# start with the full dataframe (in this case, called 'df') then..
df.sum <- df  |> 
  
  # remove missing values from continuous variables
  filter(!is.na(.data[[dep.var]])) |> 
  
  # Group the data by your categorical variable
  group_by(Group) |> 
  
  # calculate the descriptive stats
  summarize(Sample.size = n(),
            Min = min(.data[[dep.var]]),
            Q1 = quantile(.data[[dep.var]], .25),
            Median = median(.data[[dep.var]]),
            Q3 = quantile(.data[[dep.var]], .75),
            Max = max(.data[[dep.var]]),
            Mean = mean(.data[[dep.var]]),              
            SD = sd(.data[[dep.var]]))

df.sum



```

How should we round these variables? A good rule of thumb is to use the SD as the indicator of how much average variability there is in your data. Personally, I like to round the SD to 2 significant digits based on recommendations in [Cole et al. (2015)](https://www-ncbi-nlm-nih-gov.libproxy.smith.edu/pmc/articles/PMC4483789/). Then round all the other values to the same decimal place as the rounded SD.

In the case above, we should round the SD to the hundredths place to get it rounded to 2 significant digits. Thus, we will round all the values to the hundredths place.

We also don't need to show ALL these values (in fact, it may not be important to give any of these exact values if the graph itself is sufficient to get your point across). To know what values to report as well as how to analyze and graph the results, we need to see if the data has a normal distribution and if there are any outliers.

### Test for Normal Distribution

```{r}
#| label: test assumptions

#make the grouping variable a factor
df$Group = factor(df$Group,
                    levels = c("Baseline","Exercise","Recovery"))

#Test to see if each group has a normal distribution
ggpubr::ggqqplot(df, "Measurement", facet.by = "Group")


#test for outliers
df  |> 
  group_by(Group)  |> 
  rstatix::identify_outliers(Measurement)


```


The first results window to the left above shows a Q-Q plot (quantile-quantile plot) for each group comparing to a normal distribution in a graphical manner (see this [Complete Guide: How to Interpret Q-Q Plots](https://www.statology.org/qq-plot-interpretation/) if you want to learn more). Because the points are mostly close to the line and don't stray much outside of the grey shaded area, this indicates that the data are pretty normally distributed.

The last results window to the right above shows how many rows contain outliers. It says 0 rows, which indicates that there are no outliers.

### Reporting Descriptive Statistics

The results above indicate that this dataset has a normal distribution, so it would be appropriate to give the mean, SD, and number of subjects (N). The code below would round to the hundredths place and give just the values of interest in a nice formatted table (using the `flextable` package to format the table).

```{r}
#| label: formatted mean SD

#see https://davidgohel.github.io/flextable/ for more information and formatting options

#set the variable you want to summarize
dep.var <- "Measurement" 

#calculate descriptive stats
df.sum <- df |> 
  group_by(Group) |> # grouping variable
  filter(!is.na(.data[[dep.var]])) |> # remove missing values 
  
  #calculate the rounded values
  summarise(Mean = round(mean(.data[[dep.var]]), digits=2), 
            SD = signif(sd(.data[[dep.var]]), digits=2),
            N = n())


#create the formatted table
ft <- flextable(df.sum,
                cwidth = 0.75) |>  #can vary cell width as needed
  
  #bold the headings
  bold(part = "header") |> 
  
  #center columns
  align(align = "center", part = "all" )

#print the table
#right click on the table, choose select all, 
#choose copy, then paste in your document
#finish formatting as needed in your document
ft
```

If this dataset was not a normal distribution, you might choose to report the median, range (min - max), and N.

```{r}
#| label: formatted descriptive stats

#see https://davidgohel.github.io/flextable/ for more information and formatting options

#set the variable you want to summarize
dep.var <- "Measurement" 

#calculate descriptive stats
df.med <- df |> 
  group_by(Group) |> #grouping variable
  filter(!is.na(.data[[dep.var]])) |> # remove missing values 
  
  summarise(Median = round(median(.data[[dep.var]]), digits=2),
            min = round(min(.data[[dep.var]]), digits=2),
            max = round(max(.data[[dep.var]]), digits=2),
            N = n()) |> 
  mutate(Range = paste(min, max, sep = "-")) |> 
  dplyr::select(-min, -max) |> 
  relocate(N, .after = last_col())
  


#create the table
ft <- flextable(df.med,
                cwidth = 0.75) |>  #can vary cell width as needed
  
  #bold the headings
  bold(part = "header") |> 
  
  #center columns
  align(align = "center", part = "all" ) |> 
  
  set_table_properties(layout = "autofit")

#print the table
#right click on the table, choose select all, 
#choose copy, then paste in your document
#finish formatting as needed in your document
ft



```
 In general, you should choose the most appropriate way to report your descriptive statistics based on the sample size and distribution of the data. Don't give ALL the districptive statistics  - I would pick one of the options above, not both. In this case, since the data show a normal distribution, I would report the Mean, SD, and N.


## Estimation Statistics

Next, we want to perform some inferential statistical tests to see if the dependent variable was impacted by the independent variable. Since this experimental design used the same subjects in the three treatments, we need to take that into consideration using a repeated measures type of test.

First, we will use the `dabestr` package to perform a [repeated measures comparison](https://acclab.github.io/dabestr/articles/tutorial_repeated_measures.html) using estimation statistical techniques. In the example below, we will use the same baseline (compare the exercise measurements to the baseline, and also compare the recovery measurements to the baseline).

```{r}
#| label: est repeated measures

baseline_repeated_measures.mean_diff <- load(df,
  x = Group, y = Measurement,
  idx = c(
    "Baseline", "Exercise",
    "Recovery"),
  paired = "baseline", 
  id_col = ID)  |> 
  mean_diff()

print(baseline_repeated_measures.mean_diff)

dabest_plot(
  baseline_repeated_measures.mean_diff,
  swarm_label = "ln RMSSD",
  raw_marker_size = 0.5, 
  raw_marker_alpha = 0.3,
  swarm_bars = FALSE,
  contrast_bars = FALSE)

```

The top graph shows the raw dependent variable (in this case, a made up value of ln RMSSD) with a line connecting values over time for each subject (called a slopegraph, which was designed by Edward Tufte). Unfortunately, there is no way to add the mean or median for each treatment using the functions in the `dabestr` package. But it does include the sample size on the x-axis which is a plus.

The bottom graph shows the [effect size](https://www.estimationstats.com/#/about-effect-sizes) of the paired mean difference (called a Cumming estimation plot). The shaded curve shows the 5000 bootstrapped samples - basically, it is taking a random sub-sample of your data and calculating the mean difference for each, then it repeats this 5000 times (see the [Estimation Stats background material](https://www.estimationstats.com/#/background) for their explanation) and shows all these mean differences as a distribution. The black point is the mean difference between each treatment (exercise or recovery) and the baseline among all those 5000 resamples, and the dark vertical line error bars indicate the 95% confidence interval.

If you click on the other little window of results below the code chunk (the one on the left above), you will see the results of the actual statistical test.

As you can see from the Cumming estimation plot, the distribution of paired mean differences for exercise minus the baseline is well above the zero line. The mean difference is 1.2 and the lower 95% confidence interval does not go below the zero line. The p-value for a t-test conducted was less than 0.0001. This all indicates that the ln RMSSD was significatly higher during exercise than baseline. The mean ln RMSSD was only 0.30 greater on average during recovery than during the baseline recordings. The lower 95% CI does cross below zero and the p-value was 0.35. These results indicate that there was not a significant difference between the recovery period and the baseline.

## Cohen's d Effect Size

The {dabestr} package that we used above gives the raw data in a slopegraph and the estimation of the paired mean differences with a Cumming estimation plot. The package does most of the work for you, but it only gives limited ability to change the look of the set of graphs. Some people find the Cumming estimation plot difficult to interpret. 

With a little more work (and a lot more code), we can manually do a similar thing making a slope graph that also shows the points and the median or mean for each time point. We can also complete an effect test and add the results of the test to the graph. Below, we use a Cohen's d effect size for normal data with equal variance for a repeated measures design.

```{r}
#| label: cohen's d effect size

#make the grouping variable a factor with the levels ordered backwards to avoid getting negative effect sizes
df$Group = factor(df$Group,
                    levels = c("Recovery","Exercise","Baseline"))

# replace the blank below with a continuous variable of interest
dep.var <- "Measurement" 

#the code below will calculate the Cohen's d effect size when variances are unequal
    response <- dep.var
    predictor <- "Group"
    formula_string <- paste0(response, " ~ ", predictor)
    formula_object <- as.formula(formula_string)

#standardized effect size for normal data with equal variance
#Cohen's d
cohensD.results  <- 
  rstatix::cohens_d(
    data = df,         #put the dataframe name here
    formula_object,            #dependent variable ~ factor
    paired = TRUE,       #TRUE for within-subjects, FALSE for between-subjects
    var.equal = TRUE,
    hedges.correction = TRUE,
    ci = TRUE,
    conf.level = 0.95,
    ci.type = "bca",
    nboot = 1000)


#make the grouping variable a factor with levels ordered by time
df$Group = factor(df$Group,
                    levels = c("Baseline","Exercise","Recovery"))

#the code below creates a plot that connects the dots within subjects
#and also adds the Cohen's d results from the analysis above to the graph
line.plot <- df  |>    
  ggplot(aes(
    x=Group, 
    y=Measurement, 
    group = `ID`)) +
  geom_point(size = 3,  
             color = "orange", 
             alpha = 0.5) +
  geom_line(linewidth = 0.2, 
            alpha = 0.5) + 
  stat_summary(fun = mean, 
               fun.min = mean, 
               fun.max = mean, 
               geom = "crossbar", 
               color = "black",
               width = 0.2, 
               size = 0.5,
      aes(group = 'Group' )) +
  xlab("Time of Measurement") +
  ylab("ln RMSSD") +
  
   labs(title = 
          glue_data(cohensD.results[3, ], 
                    "Exercise - Baseline: Cohen's d = {signif(unname(effsize), digits=2)} (95% CI: {signif(conf.low, digits=2)} to {signif(conf.high, digits=2)})")
        ) +
  
     labs(subtitle = 
          glue_data(cohensD.results[2, ], 
                    "Recovery - Baseline: Cohen's d = {signif(unname(effsize), digits=2)} (95% CI: {signif(conf.low, digits=2)} to {signif(conf.high, digits=2)})")
        ) +
  

  #additional formatting
  theme_classic(base_size = 14)  +  #sets the font size
   theme(
    plot.title = element_text(size = 12),  # Set title size
    plot.subtitle = element_text(size = 12) # Set subtitle size to the same value
  ) +
  theme(legend.position = "none")   #controls legend/key

line.plot

```

The resulting graph is similar to the {dabestr} package slopegraph, but here we we used ggplot and have more control to manually change things. This version of the graph includes the points representing the individual subjects (orange dots) that are connected by lines within a subject as well as a black line giving the mean among subjects for each time period. We have also manually added the results of the Cohen's d effect tests above the graph. While this graph does not include an estimation plot, it does include the results of effect tests that we can more manually control which ones we want to show.

The Exercise - Baseline comparison had a large magnitude effect size of 1.8 (look at the cohensD.results dataframe and you can see how the magnitude of the effect size is categorized). The 95% Confidence interval values are also both positive, which means that the 95% CI does not cross over zero - indicating that the average ln RMSSD value was significantly greater after exercise than at baseline (you can look at the {dabestr} package Cummings estimation plot to see this shown visually). 

The Recovery - Baseline comparison had a small magnitude effect size of 0.33. The lower 95% CI was negative (below zero), but the upper 95% CI was positive (above zero). This indicates that the 95% CI crosses zero (again, you can look at the {dabestr} package Cummings estimation plot to see this shown visually) and means that the average ln RMSSD value was not significantly different from baseline.

These two different analyses gave the same general results. The only difference in how the results are displayed. You can choose which one you prefer of these two options (don't give both!).

