---
title: "Practice Data Analysis Results"
author: "Your Name Here"
date: "Sept 24, 2024"
format: 
  html:
    code-overflow: wrap
    embed-resources: true
---

## Save as

Before you get started, save this file as "dataIntro_YourName.qmd" where you replace YourName with your actual name (no spaces). This way you won't accidentally overwrite your work with a new copy of the results.qmd template.

## Load necessary packages

Run the code below at the start of the session to make sure all the necessary packages are loaded.

It is recommended that you put all your code to load packages in one R chunk at the start. Run the chunk below by clicking the little green arrow at the top right of the code chunk.

```{r}
#| label: setup
#| warning: false
#| message: false

suppressPackageStartupMessages(
  library(tidyverse)) ## for readr, dplyr, ggplot2
library(ggbeeswarm) ## for jittering points (you can remove this is you did not use a dotplot)
library(flextable) ## for making formatted tables that can be pasted into Word or Google Docs
library(dabestr) ##for data analysis using Bootstrap-coupled estimation
library(rstatix) #pipe friendly framework for basic stats
library(ggpubr) #publication ready plots
library(labelled) #for labeling variables better
```

## Import Data

Run the code chunk below to import the data.

```{r}
#| label: import data

hrvData <- read_csv("data/hrvData.csv", 
                    col_types = cols(
                      test.period = col_factor(
                        levels = c("Baseline",
                                   "Exercise", 
                                   "Recovery")), 
                      startDateTime = col_datetime(
                        format = "%Y-%m-%d %H:%M:%S")))

subjects <- read_csv("data/subjects.csv")

metadata <- read_csv("data/variable_metadata.csv")

#create variable labels using the labelled package
variable_labels <- metadata |> 
  select(variable, variable_label) |> 
  deframe()

#add labels to data
hrvData <- hrvData |> 
  left_join(subjects) |> 
  set_variable_labels(!!!variable_labels)


# for more help with labeling data see   
# https://www.pipinghotdata.com/posts/2022-09-13-the-case-for-variable-labels-in-r/    

```

## Descriptive Statistics

Calculate all the descriptive statistics for your HRV variable of choice (for example, replace "\_\_\_" with "ln.RMSSD")

```{r}
#| label: descriptive stats


# the code below will calculate descriptive statistics for a variable of interest grouped by another variable

#replace the blank with a continuous variable
x.var <- "___" 

df.sum <- hrvData  |>  
  # remove missing values from continuous variable
  filter(!is.na(.data[[x.var]])) |> 
  
  # Group the data by your categorical variable
  group_by(test.period) |> 
  
  # calculate the descriptive stats
  summarize(mean = mean(.data[[x.var]]), 
            median = median(.data[[x.var]]), 
            SD = sd(.data[[x.var]]), 
            IQR = IQR(.data[[x.var]]), 
            min = min(.data[[x.var]]),
            max = max(.data[[x.var]]),
            N = n())

df.sum

```

Now make a nice formatted table with values rounded and only the values you want to include using the script templates

Make sure to replace the blanks in the script templates with the following

-   dataframe name: hrvData

-   continuous variable name: your HRV variable of choice

-   categorical variable name: test.period

```{r}
#| label: formatted descriptive stats




```

## Statistical Results and Graph

Pick which framework you want to use to compare among treatments (estimation statistics or NHST). Then copy the code from the script templates and complete that analysis with the hrvData dataset.

Make sure to replace the blanks in the script templates with the following

-   dataframe name: hrvData

-   continuous variable name: your HRV variable of choice

-   categorical variable name: test.period

```{=html}
<!-- -->
```
-   subjects variable name: subject

```{r}
#| label: inferential stats and graph



```

## 

## Exporting your graph

How do you save a copy of your graph to use elsewhere?

The simplest thing to do is probably just right-click (Windows) or control-click (Mac) on the graph and select "Copy Image" and then just paste the image where you want it. However, if you want a higher quality image where you can specify the size, image quality, and file type you can use the `ggsave()` function.

When saving with the `ggsave()` function, you need to name the resulting file with surrounding " ", and indicate the size via height, width, and units. Don't forget to save the graph with a dpi call between 300-500 to make it nice and crisp! Look at the `ggsave()` help file for more information and options.

```{r}
#| label: Save your graph 
#| eval: false  

# save the graph! (.png or .jpg file types are recommended)
ggsave(filename = "HRVgraph.png",  
       height = 7, width = 8, units = "in", 
       dpi = 300)
```

The above code will save the last graph you made and call it "HRVgraph.png". This file will appear in the dataIntro folder under the Files tab in the lower right quadrant of this window.

If you are working on the Smith server, you will first need to export this file to your computer:

-   Check the box next to the "HRVgraph.png" file

-   Go to the "More" menu with the gear icon and click the drop down arrow

-   Select "Export" and then save the file to your computer

You can now insert this graph into your Results slide (see the lab notebook entry for today for more details).

## Render this File to HTML & Print to PDF

Now that you are all done with this activity, complete the following:

-   Save this Quarto file (make sure it has been renamed to "dataIntro_YourName.qmd")

-   Render this Quarto file to HTML (just select the Render menu item at the top of this window)

-   If the HTML file didn't already open in your browser, then click on the name of the file in the Files window and select "View in Web Browser"

-   Once the HTML file is open in your web browser, then use the print function within your browser to save it or print it as a PDF

-   Make sure the PDF file is saved somewhere on your computer that you can get to (as with the graph file above, you may first need to export this file to your computer if you are working on the Smith server)

-   Place a copy of the PDF file in your Individual Google Drive folder for this class
